A seamless, client-centric programming model for type safe web applications

### 2014-05-02 paper.tex
Javascript -> JavaScript

Development of web applications is no task for the faint of heart.
The conventional method involves splitting your program into two logical parts,
writing the one in Javascript, which is notorious even among its proponents for
being wonky and error-prone, and the other in any compiled or
server-interpreted language. Then, two are glued together using whichever
home-grown network protocol seems to fit the application. However, most web
applications are conceptually single entities, making this forced split an
undesirable hindrance which introduces new possibilities for defects, adds
development overhead and prevents code reuse.

@@

Recent advances in compiler technology from functional languages to Javascript
have led to a wealth of compilers targeting the web space, and enabled the
practical development of functional libraries and applications for
the browser. This enables us to implement our solution as a simple Haskell
library for any compiler capable of producing Javascript output, requiring no
further modification to existing compilers.

As our implementation targets the Haste Haskell to Javascript compiler, this
paper also goes into some detail about its design and implementation as well
as the alternatives available for compiling functional languages to a browser
environment.

@@

Code written in Javascript, the only widely supported language for client-side
web applications, is often confusing and error-prone, much due to the
language's lack of modularity, encapsulation facilities and type safety.

@@

  sock.onopen = function(e) {
    handshake(sock);
    msgbox.addEventListener('keydown', function(e) {
      if(e.keyCode == 13) {
        var msg = msgbox.value;
        msgbox.value = '';
        chat(msg);
      }
    });
  };
};
\end{code}
\caption{Javascript chatbox implementation}
\label{lst:javascript-client}
\end{listingfloat}

@@

After the \lstinline!export! call, we enter the domain of client-exclusive code
with the application of \lstinline!runClient!. This function executes
computations in the \lstinline!Client! monad which is essentially an
IO monad with cooperative multitasking added on top, to mitigate the fact that
Javascript has no native concurrency support. \lstinline!runClient! does not
return, and is the only function with a return type of \lstinline!App Done!,
which ensures that each \lstinline!App! computation contains exactly one client
computation.

@@

Having implemented our three server operations, all that's left is to tie them
to the client. In this tying, we see our main advantage over the Javascript
version in section \ref{sec:jsexample} in action: the \lstinline!export!
function builds a strongly typed bridge between the client and the server,
ensuring that any future enhancements to our chatbox program are made safely,
in one place, instead of being spread about throughout two disjoint code bases.

@@

The \lstinline!recvLoop! function perpetually asks the server for new messages
and updates the chat log whenever one arrives. Note that unlike the
\lstinline!onmessage! callback of the Javascript version of this example,
\lstinline!recvLoop! is acting as a completely self-contained process with
linear program flow, keeping track of its own state and only reaching out to
the outside world to write its state to the chat log whenever necessary. As
the \lstinline!awaitMsg! function blocks until a message arrives,
\lstinline!recvLoop! will make exactly one iteration per received message.

@@

\paragraph{Two compilers}
The principal trick to our solution is compiling the same program twice; once
with a compiler that generates the server binary, and once with one that
generates Javascript. Conditional compilation is used for a select few
functions, to enable slightly different behavior on the client and on the
server as necessary. As we believe that a monadic programming style is a good
fit for this application, we have chosen to use Haskell as the base of our
solution, which by default leads us to choose GHC as our server-side compiler.
For reasons outlined in section \ref{sec:haste}, the Haste compiler was chosen
to provide the client-side code.

@@

\paragraph{The \lstinline!Client! monad and the \lstinline!onServer! function}
As synchronous network communication is one of our stated goals, it is clear
that we will need some kind of blocking primitive. Since Javascript does not
support any kind of blocking, we will have to implement this ourselves.

A solution is given in the \emph{poor man's concurrency
monad}\ \cite{concurrencymonad}. Making use of a continuation monad with
primitive operations for forking a computation and atomically lifting an IO
computation into the monad, it is possible to implement cooperative
multitasking on top of the non-concurrent Javascript runtime. This monad
allows us to implement \lstinline!MVar!s as our blocking primitive, with the
same semantics as their regular Haskell counterpart.\ \cite{ffi}
This concurrency-enhanced monad is used as the basis of the \lstinline!Client!
monad.

@@

In order to allow the same language to be used on both client and server, we
need some way to compile that language into Javascript. To this end, we make
use of the Haste compiler\ \cite{haste}, started as an MSc thesis and continued
as part of this work. Haste builds on the GHC compiler to provide the full
Haskell language, including most GHC-specific extensions, in the browser.

@@

Haste is by no means the only Javascript-targeting compiler for a purely
functional language. In particular, the GHC-based GHCJS\ \cite{ghcjs} and
UHC\ \cite{uhc} compilers are both capable of compiling standard Haskell into
Javascript; the Fay\ \cite{fay} language was designed from the ground up to
target the web space using a subset of Haskell; and there exist solutions for
compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to Javascript as
well.

@@

\begin{itemize}
  \item The language must provide a static type system, since one of our
        primary concerns is to reduce defect rates through static typing of
        the client-server communication channel.
  \item The language must be compilable to both Javascript and a format
        suitable for server-side execution as we want our web applications
        to be written and compiled as a single program.
  \item We want the language to provide decent support for a monadic
        programming style, as we our abstractions for cooperative multitasking
        and synchronous client-server communication are neatly expressible in
        this style.
\end{itemize}

@@

One such tradeoff is the decision to make use of the facilities provided by the
web browser - most prominently the Javascript engine's native garbage
collector - as far as possible rather than implementing parts of the GHC
runtime in Javascript. In the same vein, a conscious effort is made to generate
as ``regular-looking'' code as possible. This results in large gains in code
compactness and execution speed - in our experiences, compiler Haste programs
are often an order of magnitude smaller and several times faster than an
equivalent program compiled with some other Haskell to Javascript compiler.

In some rare cases, however, this means having to sacrifice compatibility with
low-level GHC features. Most notably, Haste does not support weak references
or pre-emptive multitasking, as these would require modifications to the
runtime system and generated code which confound the structure of the
resulting Javascript and have a detrimental impact on code size and execution
speed.

@@

\begin{itemize}
  \item STG is still a functional intermediate representation, based on the
        lambda calculus. When generating code for a high level target language
        such as Javascript, where functions are first class objects, this
        allows for a higher level translation than when doing traditional
        compilation to lower level targets like stack machines or register
        machines. This in turn allows us to make more efficient use of the
        target language's runtime, leading to smaller, faster code.
  \item In contrast to Haskell itself and GHC's intermediate Core language, STG
        represents `thunks`, the construct used by GHC to implement non-strict
        evaluation, as closures which are explicitly created and evaluated.
        Closures are decorated with a wealth of information, such as their set
        of captured varibles, any type information needed for code generation,
        and so on. While extracting this information manually is not very hard,
        having this done for us means we can get away with a simpler
        compilation pipeline.
  \item The language is very small, essentially only comprising lambda
        abstraction and application, plus primitive operations and facilities
        for calling out to other languages. Again, this allows the Haste
        compiler to be a very simple thing indeed.
  \item Any extensions to the Haskell language implemented by GHC will already
        have been translated into this very simple intermediate format,
        allowing us to support basically any extension GHC supports without
        effort.
  \item Application of external functions is always saturated, as is
        application of most other functions. This allows for compiling most
        function applications into simple Javascript function calls, limiting
        the use of the slower dynamic techniques required to handle curried
        functions in the general case\ \cite{fastcurry} to cases where it is
        simply not possible to statically determine the arity of a function.
\end{itemize}

@@

The runtime data representation of Haste programs is kept as close to regular
Javascript programs as possible. The numeric types are represented using the
Javascript \lstinline!Number! type, which is defined as the IEEE754 double
precision floating point type. This adds some overhead to operations on
integers as overflow and non-integer divisions must be handled. However, this
is common practice in hand-written Javascript as well, and is generally handled
efficiently by Javascript engines.

@@

Functions are represented as plain Javascript functions, one of the blessings
of targeting a high level language, and application can therefore be
implemented as its Javascript counterpart in most cases. In the general case,
however, functions may be curried. For such cases where the arity of an applied
function can not be determined statically, application is implemented using the
eval/apply method described in\ \cite{fastcurry} instead.

\subsection{Interfacing with Javascript}

While Haste supports the Foreign Function Interface inherited from GHC, with
its usual features and limitations \ \cite{ffi}, it is often impractical to
work within the confines of an interface designed for communication on a very
low level. For this reason Haste sports its own method for interacting with
Javascript as well, which allows the programmer to pass any value back and
forth between Haskell and Javascript, as long as she can come up with a way to
translate this value between its Haskell and Javascript representations. Not
performing any translation at all is also a valid ``translation'', which allows
Haskell code to store any Javascript value for later retrieval without
inspecting it and vice versa. The example given in figure \ref{lst:ffi}
implements mutable variables using this custom Javascript interface.

@@

-- A MutableVar is completely opaque to Haskell code
-- and is only ever manipulated in Javascript. Thus,
-- we use the Unpacked type to represent it,
-- indicating a completely opaque value.
newtype MutableVar a = MV Unpacked

@@

The core of this interface consists of the \lstinline!ffi! function, which
allows the programmer to create a Haskell function from arbitrary Javascript
code. This function exploits Javascript's ability to parse and execute
arbitrary strings at run time using the \lstinline!eval! function, coupled with
the fact that functions in Haste and in Javascript share the same
representation, to dynamically create a function object at runtime.
The \lstinline!ffi! function is typed using the same method as the
\lstinline!mkExported! function described in section \ref{sec:impl}.
When applied to one or more arguments instantiating the \lstinline!Marshal!
type class, the \lstinline!pack! function is applied to each argument,
marshalling them into their respective Javascript representations, before they
are passed to the dynamically created function. When that function returns,
the inverse \lstinline!unpack! function is applied to its return value before
it is passed back into the Haskell world.

@@

There are several benefits to this method, the most prominent being that new
marshallable types can be added by simply instantiating a type class. Thanks
to the lazy evaluation employed by Haste, each foreign function object is only
created once and then cached; any further calls to the same (Haskell) function
will reuse the cached function object. Implementation-wise, this method is also
very non-intrusive, requiring only the use of the normal FFI to import
Javascript's \lstinline!eval! function; no modification of the compiler is
needed.

@@

\paragraph{Conductance} Conductance is an application server built on
StratifiedJS, a Javascript language extension which adds a few niceties such as
cooperative multitasking and more concise syntax for many common tasks.
Conductance uses an RPC-based model for client-server communication, much like
our own, but also adds the possibility for the server to independently transmit
data back to the client through the use of shared variables or call back into
the client by way of function objects received via RPC call, as well as the
possibility for both client and server to seamlessly modify variables located
on the opposite end of the network. Conductance is quite new and has no
relevant publications. It is, however, used for several large scale web
applications.

While Conductance gets rid of the callback-based programming model endemic to
regular Javascript, it still suffers from many of its usual drawbacks. In
particular, the weak typing of Javascript poses a problem in that the
programmer is in no way reprimanded by her tools for using server APIs
incorrectly or trying to transmit values which can not be sensibly serialized
and de-serialized, such as DOM nodes. Wrongly typed programs will thus crash, or
even worse, gleefully keep running with erroneous state due to implicit type
conversions, rather than give the programmer some advance warning that something
is amiss.

@@

Like Conductance, Duetto suffers somewhat from its heritage: while the
client-side code is not memory-unsafe, as it is not possible to generate
memory-unsafe Javascript code, its server-side counterpart unfortunately is.
Our reservations expressed about how network communication in Duetto can be
initiated implicitly apply to Duetto as well.

\paragraph{Sunroof} In contrast to Conductance and Duetto, Sunroof is an
embedded language. Implemented as a Haskell library, it allows the programmer
to use Haskell to write code which is compiled to Javascript and executed on
the client. The language can best be described as having Javascript semantics
with Haskell's type system. Communication between client and server is
accomplished through the use of ``downlinks'' and ``uplinks'', allowing for
data to be sent to and from the client respectively.

Sunroof is completely type-safe, in the DSL itself as well as in the
communication with the Haskell host. However, the fact that client and server
must be written in two separate languages - any code used to generate
Javascript must be built solely from the primitives of the Sunroof
language in order to be compilable into Javascript, precluding use of general
Haskell code - makes code reuse hard.

@@

Our programming model is implemented as a library, assuming only two Haskell
compilers, one targeting Javascript and one targeting the programmer's server
platform of choice. While we use Haste as our Javascript-targeting compiler,
modifying our implementation to use GHCJS or even the Javascript backend of UHC
would be trivial. This implementation not only allows for greater flexibility,
but also eliminates the need to tangle with complex compiler internals.

@@

This kind of divergence can be worked around using conditional compilation, but
is highly problematic even so; using a unified package database between the two
compilers, while problematic due to the differing natures of native and
Javascript compilation respectively, would be a significant improvement in this
area.

@@

\bibitem[Guthrie (2014)]{jserlang}
G. Guthrie. (2014, January 1). "Your transpiler to Javascript toolbox". Retrieved March 1, 2014, from http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/.

@@

\bibitem[Nazarov (2012)]{ghcjs}
V. Nazarov. "GHCJS Haskell to Javascript Compiler". Retrieved March 1, 2014, from https://github.com/ghcjs/ghcjs.

>>>>

Development of web applications is no task for the faint of heart.
The conventional method involves splitting your program into two logical parts,
writing the one in JavaScript, which is notorious even among its proponents for
being wonky and error-prone, and the other in any compiled or
server-interpreted language. Then, two are glued together using whichever
home-grown network protocol seems to fit the application. However, most web
applications are conceptually single entities, making this forced split an
undesirable hindrance which introduces new possibilities for defects, adds
development overhead and prevents code reuse.

@@

Recent advances in compiler technology from functional languages to JavaScript
have led to a wealth of compilers targeting the web space, and enabled the
practical development of functional libraries and applications for
the browser. This enables us to implement our solution as a simple Haskell
library for any compiler capable of producing JavaScript output, requiring no
further modification to existing compilers.

As our implementation targets the Haste Haskell to JavaScript compiler, this
paper also goes into some detail about its design and implementation as well
as the alternatives available for compiling functional languages to a browser
environment.

@@

Code written in JavaScript, the only widely supported language for client-side
web applications, is often confusing and error-prone, much due to the
language's lack of modularity, encapsulation facilities and type safety.

@@

  sock.onopen = function(e) {
    handshake(sock);
    msgbox.addEventListener('keydown', function(e) {
      if(e.keyCode == 13) {
        var msg = msgbox.value;
        msgbox.value = '';
        chat(msg);
      }
    });
  };
};
\end{code}
\caption{JavaScript chatbox implementation}
\label{lst:javascript-client}
\end{listingfloat}

@@

After the \lstinline!export! call, we enter the domain of client-exclusive code
with the application of \lstinline!runClient!. This function executes
computations in the \lstinline!Client! monad which is essentially an
IO monad with cooperative multitasking added on top, to mitigate the fact that
JavaScript has no native concurrency support. \lstinline!runClient! does not
return, and is the only function with a return type of \lstinline!App Done!,
which ensures that each \lstinline!App! computation contains exactly one client
computation.

@@

Having implemented our three server operations, all that's left is to tie them
to the client. In this tying, we see our main advantage over the JavaScript
version in section \ref{sec:jsexample} in action: the \lstinline!export!
function builds a strongly typed bridge between the client and the server,
ensuring that any future enhancements to our chatbox program are made safely,
in one place, instead of being spread about throughout two disjoint code bases.

@@

The \lstinline!recvLoop! function perpetually asks the server for new messages
and updates the chat log whenever one arrives. Note that unlike the
\lstinline!onmessage! callback of the JavaScript version of this example,
\lstinline!recvLoop! is acting as a completely self-contained process with
linear program flow, keeping track of its own state and only reaching out to
the outside world to write its state to the chat log whenever necessary. As
the \lstinline!awaitMsg! function blocks until a message arrives,
\lstinline!recvLoop! will make exactly one iteration per received message.

@@

\paragraph{Two compilers}
The principal trick to our solution is compiling the same program twice; once
with a compiler that generates the server binary, and once with one that
generates JavaScript. Conditional compilation is used for a select few
functions, to enable slightly different behavior on the client and on the
server as necessary. As we believe that a monadic programming style is a good
fit for this application, we have chosen to use Haskell as the base of our
solution, which by default leads us to choose GHC as our server-side compiler.
For reasons outlined in section \ref{sec:haste}, the Haste compiler was chosen
to provide the client-side code.

@@

\paragraph{The \lstinline!Client! monad and the \lstinline!onServer! function}
As synchronous network communication is one of our stated goals, it is clear
that we will need some kind of blocking primitive. Since JavaScript does not
support any kind of blocking, we will have to implement this ourselves.

A solution is given in the \emph{poor man's concurrency
monad}\ \cite{concurrencymonad}. Making use of a continuation monad with
primitive operations for forking a computation and atomically lifting an IO
computation into the monad, it is possible to implement cooperative
multitasking on top of the non-concurrent JavaScript runtime. This monad
allows us to implement \lstinline!MVar!s as our blocking primitive, with the
same semantics as their regular Haskell counterpart.\ \cite{ffi}
This concurrency-enhanced monad is used as the basis of the \lstinline!Client!
monad.

@@

In order to allow the same language to be used on both client and server, we
need some way to compile that language into JavaScript. To this end, we make
use of the Haste compiler\ \cite{haste}, started as an MSc thesis and continued
as part of this work. Haste builds on the GHC compiler to provide the full
Haskell language, including most GHC-specific extensions, in the browser.

@@

Haste is by no means the only JavaScript-targeting compiler for a purely
functional language. In particular, the GHC-based GHCJS\ \cite{ghcjs} and
UHC\ \cite{uhc} compilers are both capable of compiling standard Haskell into
JavaScript; the Fay\ \cite{fay} language was designed from the ground up to
target the web space using a subset of Haskell; and there exist solutions for
compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to JavaScript as
well.

@@

\begin{itemize}
  \item The language must provide a static type system, since one of our
        primary concerns is to reduce defect rates through static typing of
        the client-server communication channel.
  \item The language must be compilable to both JavaScript and a format
        suitable for server-side execution as we want our web applications
        to be written and compiled as a single program.
  \item We want the language to provide decent support for a monadic
        programming style, as we our abstractions for cooperative multitasking
        and synchronous client-server communication are neatly expressible in
        this style.
\end{itemize}

@@

One such tradeoff is the decision to make use of the facilities provided by the
web browser - most prominently the JavaScript engine's native garbage
collector - as far as possible rather than implementing parts of the GHC
runtime in JavaScript. In the same vein, a conscious effort is made to generate
as ``regular-looking'' code as possible. This results in large gains in code
compactness and execution speed - in our experiences, compiler Haste programs
are often an order of magnitude smaller and several times faster than an
equivalent program compiled with some other Haskell to JavaScript compiler.

In some rare cases, however, this means having to sacrifice compatibility with
low-level GHC features. Most notably, Haste does not support weak references
or pre-emptive multitasking, as these would require modifications to the
runtime system and generated code which confound the structure of the
resulting JavaScript and have a detrimental impact on code size and execution
speed.

@@

\begin{itemize}
  \item STG is still a functional intermediate representation, based on the
        lambda calculus. When generating code for a high level target language
        such as JavaScript, where functions are first class objects, this
        allows for a higher level translation than when doing traditional
        compilation to lower level targets like stack machines or register
        machines. This in turn allows us to make more efficient use of the
        target language's runtime, leading to smaller, faster code.
  \item In contrast to Haskell itself and GHC's intermediate Core language, STG
        represents `thunks`, the construct used by GHC to implement non-strict
        evaluation, as closures which are explicitly created and evaluated.
        Closures are decorated with a wealth of information, such as their set
        of captured varibles, any type information needed for code generation,
        and so on. While extracting this information manually is not very hard,
        having this done for us means we can get away with a simpler
        compilation pipeline.
  \item The language is very small, essentially only comprising lambda
        abstraction and application, plus primitive operations and facilities
        for calling out to other languages. Again, this allows the Haste
        compiler to be a very simple thing indeed.
  \item Any extensions to the Haskell language implemented by GHC will already
        have been translated into this very simple intermediate format,
        allowing us to support basically any extension GHC supports without
        effort.
  \item Application of external functions is always saturated, as is
        application of most other functions. This allows for compiling most
        function applications into simple JavaScript function calls, limiting
        the use of the slower dynamic techniques required to handle curried
        functions in the general case\ \cite{fastcurry} to cases where it is
        simply not possible to statically determine the arity of a function.
\end{itemize}

@@

The runtime data representation of Haste programs is kept as close to regular
JavaScript programs as possible. The numeric types are represented using the
JavaScript \lstinline!Number! type, which is defined as the IEEE754 double
precision floating point type. This adds some overhead to operations on
integers as overflow and non-integer divisions must be handled. However, this
is common practice in hand-written JavaScript as well, and is generally handled
efficiently by JavaScript engines.

@@

Functions are represented as plain JavaScript functions, one of the blessings
of targeting a high level language, and application can therefore be
implemented as its JavaScript counterpart in most cases. In the general case,
however, functions may be curried. For such cases where the arity of an applied
function can not be determined statically, application is implemented using the
eval/apply method described in\ \cite{fastcurry} instead.

\subsection{Interfacing with JavaScript}

While Haste supports the Foreign Function Interface inherited from GHC, with
its usual features and limitations \ \cite{ffi}, it is often impractical to
work within the confines of an interface designed for communication on a very
low level. For this reason Haste sports its own method for interacting with
JavaScript as well, which allows the programmer to pass any value back and
forth between Haskell and JavaScript, as long as she can come up with a way to
translate this value between its Haskell and JavaScript representations. Not
performing any translation at all is also a valid ``translation'', which allows
Haskell code to store any JavaScript value for later retrieval without
inspecting it and vice versa. The example given in figure \ref{lst:ffi}
implements mutable variables using this custom JavaScript interface.

@@

-- A MutableVar is completely opaque to Haskell code
-- and is only ever manipulated in JavaScript. Thus,
-- we use the Unpacked type to represent it,
-- indicating a completely opaque value.
newtype MutableVar a = MV Unpacked

@@

The core of this interface consists of the \lstinline!ffi! function, which
allows the programmer to create a Haskell function from arbitrary JavaScript
code. This function exploits JavaScript's ability to parse and execute
arbitrary strings at run time using the \lstinline!eval! function, coupled with
the fact that functions in Haste and in JavaScript share the same
representation, to dynamically create a function object at runtime.
The \lstinline!ffi! function is typed using the same method as the
\lstinline!mkExported! function described in section \ref{sec:impl}.
When applied to one or more arguments instantiating the \lstinline!Marshal!
type class, the \lstinline!pack! function is applied to each argument,
marshalling them into their respective JavaScript representations, before they
are passed to the dynamically created function. When that function returns,
the inverse \lstinline!unpack! function is applied to its return value before
it is passed back into the Haskell world.

@@

There are several benefits to this method, the most prominent being that new
marshallable types can be added by simply instantiating a type class. Thanks
to the lazy evaluation employed by Haste, each foreign function object is only
created once and then cached; any further calls to the same (Haskell) function
will reuse the cached function object. Implementation-wise, this method is also
very non-intrusive, requiring only the use of the normal FFI to import
JavaScript's \lstinline!eval! function; no modification of the compiler is
needed.

@@

\paragraph{Conductance} Conductance is an application server built on
StratifiedJS, a JavaScript language extension which adds a few niceties such as
cooperative multitasking and more concise syntax for many common tasks.
Conductance uses an RPC-based model for client-server communication, much like
our own, but also adds the possibility for the server to independently transmit
data back to the client through the use of shared variables or call back into
the client by way of function objects received via RPC call, as well as the
possibility for both client and server to seamlessly modify variables located
on the opposite end of the network. Conductance is quite new and has no
relevant publications. It is, however, used for several large scale web
applications.

While Conductance gets rid of the callback-based programming model endemic to
regular JavaScript, it still suffers from many of its usual drawbacks. In
particular, the weak typing of JavaScript poses a problem in that the
programmer is in no way reprimanded by her tools for using server APIs
incorrectly or trying to transmit values which can not be sensibly serialized
and de-serialized, such as DOM nodes. Wrongly typed programs will thus crash, or
even worse, gleefully keep running with erroneous state due to implicit type
conversions, rather than give the programmer some advance warning that something
is amiss.

@@

Like Conductance, Duetto suffers somewhat from its heritage: while the
client-side code is not memory-unsafe, as it is not possible to generate
memory-unsafe JavaScript code, its server-side counterpart unfortunately is.
Our reservations expressed about how network communication in Duetto can be
initiated implicitly apply to Duetto as well.

\paragraph{Sunroof} In contrast to Conductance and Duetto, Sunroof is an
embedded language. Implemented as a Haskell library, it allows the programmer
to use Haskell to write code which is compiled to JavaScript and executed on
the client. The language can best be described as having JavaScript semantics
with Haskell's type system. Communication between client and server is
accomplished through the use of ``downlinks'' and ``uplinks'', allowing for
data to be sent to and from the client respectively.

Sunroof is completely type-safe, in the DSL itself as well as in the
communication with the Haskell host. However, the fact that client and server
must be written in two separate languages - any code used to generate
JavaScript must be built solely from the primitives of the Sunroof
language in order to be compilable into JavaScript, precluding use of general
Haskell code - makes code reuse hard.

@@

Our programming model is implemented as a library, assuming only two Haskell
compilers, one targeting JavaScript and one targeting the programmer's server
platform of choice. While we use Haste as our JavaScript-targeting compiler,
modifying our implementation to use GHCJS or even the JavaScript backend of UHC
would be trivial. This implementation not only allows for greater flexibility,
but also eliminates the need to tangle with complex compiler internals.

@@

This kind of divergence can be worked around using conditional compilation, but
is highly problematic even so; using a unified package database between the two
compilers, while problematic due to the differing natures of native and
JavaScript compilation respectively, would be a significant improvement in this
area.

@@

\bibitem[Guthrie (2014)]{jserlang}
G. Guthrie. (2014, January 1). "Your transpiler to JavaScript toolbox". Retrieved March 1, 2014, from http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/.

@@

\bibitem[Nazarov (2012)]{ghcjs}
V. Nazarov. "GHCJS Haskell to JavaScript Compiler". Retrieved March 1, 2014, from https://github.com/ghcjs/ghcjs.
